Fractional Decomposition Tree Algorithm: A tool for studying the integrality gap of Integer Programs

### 2020-08-10 FDT2020/main-section.tex
Responding to Arash's comments.

In this section we prove Theorem \ref{domtoIP}. In fact, we prove a stronger result. 
\begin{lemma}\label{domlemma}
	Given an integral vector in the dominant of an LP relaxation ($\tilde{x}\in \dom(P)$ and $\tilde{x}\in \{0,1\}^n$) for an IP instance that has integrality gap $g < \infty$, there is an algorithm (the DomToIP algorithm) that finds $\bar{x}\in S$ in polynomial time such that $\bar{x}\leq \tilde{x}$.\end{lemma}
Lemma \ref{domlemma} implies Theorem \ref{domtoIP}, since it is easy to obtain an integer point in $\dom(P)$: numerically rounding up $x^*$ (or any fractional point in $P$) gives us a point in $\dom(P)$.
\cindy{I added the caveat that the instance has a finite integrality gap and made it a little more self-contained.  Is it OK?}
Hence, we can assume in the proof below that $\tilde{x}_i= 0$ for $i=t+1,\ldots,n$. 
%\arash{I also made a change based on the change above}\cindy{Thanks!}
\subsection{Proof of Lemma \ref{domlemma}: The DomToIP Algorithm}

We introduce an algorithm that considers each of the variables iteratively, starting from the first coordinate and ending at the $t$-th coordinate. If the initial (integral) point $\tilde{x}$ has the variable set to $1$, the algorithm tries to reduce the variable to $0$ if possible. \arash{A caution here: the phrase if possible a bit vague here. We round down to zero, if there is feasible solution to LP (so a point in $P(I)$, such that the first $\ell$ coordinates of it are all equal to the point $x^{(\ell)}$ that we have in iteration $\ell$ and its $\ell+1$-th coordinate is 0}. We run the algorithm for $\ell\in \{0,\ldots,t-1\}$ iterations and in each iteration we find $x^{(\ell)}\in \dom(P)$  such that $x^{(\ell)}_i\in \{0,1\}$ for $i=1,\ldots,\ell$.
\cindy{I think the previous sentence also added to my confusion before (along with the term ``fix'' which I have replaced).  After $\ell$ iterations the first $\ell$ components are integer, but they were integer before.  Did you intend to say that we finalize a 0-1 value for the first $\ell$ coordinates?}
We can set $x^{(0)}=\tilde{x}$. Now consider the following linear program. The variables of this linear program are the $z\in \mathbb{R}^n$.
\begin{align}
\HelperLP(x^{(\ell)})\quad\quad& \min\quad \;z_{\ell+1}\\
&\;\text{s.t.} \quad \;\;Az\geq b \\
&\;{\color{white}{\text{s.t.}} }\quad \;\; \; z_j = x^{(\ell)}_j \quad \; j =1,\ldots, \ell\\
&\;{\color{white}{\text{s.t.}} }\quad \; \;\; z_j \leq x^{(\ell)}_j \quad \; j = \ell+1,\ldots,n\\
&\;{\color{white}{\text{s.t.}} }\quad \; \;\; z\;\geq 0
\end{align}

If the optimal value to $\HelperLP(x^{(\ell)})$ is 0, then let $x^{(\ell+1)}_{\ell+1} = 0$. Otherwise if the optimal value is strictly positive let $x^{(\ell+1)}_{\ell+1} = 1$. Let $x^{(\ell+1)}_j = x^{(\ell)}_j$ for $j\in [n]\setminus \{\ell+1\}$ (See Algorithm \ref{domtoIPalg}).

The above procedure shows how to find $x^{(\ell+1)}$ from $x^{(\ell)}$. The DomToIP algorithm initializes with $x^{(0)}=\tilde{x}$ and  iteratively calls this procedure in order to obtain $x^{(t)}$.

@@

	Let $x^0$, $x^1$, $\lambda_0$, and $\lambda_1$ be an optimal solution to the LP above. Set $\gamma_0=\lambda_0$ and $\gamma_1 = \lambda_1$.
\cindy{The sentence right before this is how I think you link the $\lambda$ in the LP to the $\gamma$ constants in the Lemma statement and the claim below, etc.}
Let $\hat{x}^0 = x^0/\gamma_0$, $\hat{x}^1=x^1/\gamma_1$. This choice satisfies  (ii), (iii), (iv). To show that (i) is also satisfied we prove the following claim.
	\arash{Now that I notice, we use $\lambda_0,\lambda_1,x^0,x^1}$ to represnt a feasible solution to the system 13-18. Since the optimal solution is use only for the duration of one sentence above maybe we can change the notation for the optimal solution. For instance use $y^0,y^1, \lambda^*_0,\lambda^*_1$}
	\begin{claim}\label{CVexists}
		We have $\gamma_0 + \gamma_1\geq 1/g$.
	\end{claim}
	\begin{cproof}
		We show that there is a feasible solution that achieves the objective value of $\frac{1}{g}$. By Theorem \ref{CV2} there exists $\theta \in [0,1]^k$, with $\sum_{i=1}^{k}\theta_i = 1$ and $\tilde{x}^i\in S$ for $i\in[k]$ such that 
		$\sum_{i=1}^{k}\theta_i \tilde{x}^i\leq gx'$. So

>>>>

In this section we prove Theorem \ref{domtoIP}. In fact, we prove a stronger result. 
\begin{lemma}\label{domlemma}
	Given an integral vector in the dominant of an LP relaxation ($\tilde{x}\in \dom(P)$ and $\tilde{x}\in \{0,1\}^n$) for an IP instance that has integrality gap $g < \infty$, there is an algorithm (the DomToIP algorithm) that finds $\bar{x}\in S$ in polynomial time such that $\bar{x}\leq \tilde{x}$.\end{lemma}
Lemma \ref{domlemma} implies Theorem \ref{domtoIP}, since it is easy to obtain an integer point in $\dom(P)$: numerically rounding up $x^*$ (or any fractional point in $P$) gives us a point in $\dom(P)$.
%\cindy{I added the caveat that the instance has a finite integrality gap and made it a little more self-contained.  Is it OK?}
Hence, we can assume in the proof below that $\tilde{x}_i= 0$ for $i=t+1,\ldots,n$. 
%\arash{I also made a change based on the change above}\cindy{Thanks!}
\subsection{Proof of Lemma \ref{domlemma}: The DomToIP Algorithm}

We introduce an algorithm that builds a solution from the input integral point $\tilde{x}$. It finalizes a binary value for each variable iteratively, starting from the first coordinate and ending at the $t$-th coordinate.
%Any coordinate that is $0$ in $\tilde{x}$ remains $0$ in the solution.
In iteration $\ell\in \{0,\ldots,t-1\}$, if $\tilde{x}_{\ell+1}=1$, the algorithm reduces the $\ell + 1$st coordinate of the solution to $0$ if there is an LP-feasible point $x'$ that respects the decisions made so far, and has $x'_{\ell}=0$. Intuitively, we can fix this coordinate to $0$ in the partial solution and still set the remaining coordinates to find the solution promised in Lemma~\ref{domlemma}.
\arash{A caution here: the phrase if possible a bit vague here. We round down to zero, if there is feasible solution to LP (so a point in $P(I)$, such that the first $\ell$ coordinates of it are all equal to the point $x^{(\ell)}$ that we have in iteration $\ell$ and its $\ell+1$-th coordinate is 0}.
\cindy{Yes.  I was just trying to give intuition. In particular, I was trying to distinguish Dom2IP that starts with an integral vector, where we push $1$s down, from FDT, where we start with a fractional vector and progressively set components to binary values. Do you think the above version is OK?}
More specifically, in iteration $\ell\in \{0,\ldots,t-1\}$ we produce $x^{(\ell)}\in \dom(P)$ such that $x^{(\ell)}_i\in \{0,1\}$ for $i=1,\ldots,\ell$. These $\ell$ components will not change for the remainder of the algorithm.
%We run the algorithm for $\ell\in \{0,\ldots,t-1\}$ iterations and in each iteration we find $x^{(\ell)}\in \dom(P)$  such that $x^{(\ell)}_i\in \{0,1\}$ for $i=1,\ldots,\ell$.
%\cindy{I think the previous sentence also added to my confusion before (along with the term ``fix'' which I have replaced).  After $\ell$ iterations the first $\ell$ components are integer, but they were integer before.  Did you intend to say that we finalize a 0-1 value for the first $\ell$ coordinates?}

We can set $x^{(0)}=\tilde{x}$. We now show how to find $x^{(\ell+1)}$ from $x^{(\ell)}$. Consider the following linear program. The variables of this linear program are the $z\in \mathbb{R}^n$.
\begin{align}
\HelperLP(x^{(\ell)})\quad\quad& \min\quad \;z_{\ell+1}\\
&\;\text{s.t.} \quad \;\;Az\geq b \\
&\;{\color{white}{\text{s.t.}} }\quad \;\; \; z_j = x^{(\ell)}_j \quad \; j =1,\ldots, \ell\\
&\;{\color{white}{\text{s.t.}} }\quad \; \;\; z_j \leq x^{(\ell)}_j \quad \; j = \ell+1,\ldots,n\\
&\;{\color{white}{\text{s.t.}} }\quad \; \;\; z\;\geq 0
\end{align}

If the optimal value to $\HelperLP(x^{(\ell)})$ is 0, then let $x^{(\ell+1)}_{\ell+1} = 0$. Otherwise if the optimal value is strictly positive let $x^{(\ell+1)}_{\ell+1} = 1$. Let $x^{(\ell+1)}_j = x^{(\ell)}_j$ for $j\in [n]\setminus \{\ell+1\}$. The DomToIP algorithm initializes with $x^{(0)}=\tilde{x}$ and iteratively calls this procedure in order to obtain $x^{(t)}$ (See Algorithm \ref{domtoIPalg}).
\cindy{I rearranged the above paragraph a bit without changing the content.  Please see if you think it's OK.}

@@

	Let $y^0$, $y^1$, $\lambda^*_0$, and $\lambda^*_1$ be an optimal solution to the LP above. Set $\gamma_0=\lambda^*_0$ and $\gamma_1 = \lambda^*_1$.
%Let $x^0$, $x^1$, $\lambda_0$, and $\lambda_1$ be an optimal solution to the LP above. Set $\gamma_0=\lambda_0$ and $\gamma_1 = \lambda_1$.
\cindy{The sentence right before this is how I think you link the $\lambda$ in the LP to the $\gamma$ constants in the Lemma statement and the claim below, etc.}
Let $\hat{x}^0 = y^0/\lambda^*_0$, $\hat{x}^1=y^1/\lambda^*_1$. This choice satisfies  (ii), (iii), (iv). To show that (i) is also satisfied we prove the following claim.
	\arash{Now that I notice, we use $\lambda_0,\lambda_1,x^0,x^1$ to represnt a feasible solution to the system 13-18. Since the optimal solution is use only for the duration of one sentence above maybe we can change the notation for the optimal solution. For instance use $y^0,y^1, \lambda^*_0,\lambda^*_1$}
\cindy{I think I made the changes you suggested.  I also used the LP optima to set the $\hat{x}$.  Please check.}
	\begin{claim}\label{CVexists}
		We have $\gamma_0 + \gamma_1\geq 1/g$.
	\end{claim}
	\begin{cproof}
		We show that there is a feasible solution that achieves the objective value of $\frac{1}{g}$. By Theorem \ref{CV2} there exists $\theta \in [0,1]^k$, with $\sum_{i=1}^{k}\theta_i = 1$ and $\tilde{x}^i\in S$ for $i\in[k]$ such that 
		$\sum_{i=1}^{k}\theta_i \tilde{x}^i\leq gx'$. So
